---
title: "SA Species Sensitivity"
author: "Kathryn Sokolowski"
date: '2022-06-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load packages
```{r}
library(here)
library(viridis)
library(randomForest)
library(colorspace)
library(gam)
library(viridis) #the best colors
library(scales) #alpha transparency
library(mgcv) #Wood's implementation of generalized additive models
library(randomForest)
library(tidyverse)
library(dplyr)
library(tibble)
library(ggforce)
```

Dataset has been cut down from full to only include Southern African countries. Sooo while this might be a bad idea, it at least eliminates huge equatorial geography problems. 

Using Hempson's models of herbivore biomass we take that model to turn it into present vs absent. Those present/absent values are tied to 0.5 degree sections across South Africa. Then these are matched to those corresponding worldclim values. So instead of being based around the specific locality, it is modeled across all of South Africa. Also add aridity in. 

```{r}
Hempson <- here("Data", "Hempson-Venter-Updated.csv") %>% 
  read_csv()  %>%
  na.omit() 

min(Hempson$bio1)
max(Hempson$bio1)
min(Hempson$mean_wc)
max(Hempson$mean_wc)
min(Hempson$bio15)
max(Hempson$bio15)

as_factor(Hempson)
```

Check for correlates
```{r}
env <- here("Data", "southafrica-wide-env.csv") %>% 
  na.omit() %>%
  read_csv()

corenvironment <- env %>%
  select(Longitude, Latitude, mean_wc, AridityIndex, Bio1, Bio6, Bio15) %>%
  na.omit()
cor(corenvironment)
```

Do random forest models with updated Hempson data
```{r}

Hempson <- Hempson %>% 
  select(Latitude, Longitude, mean_wc, AridityIndex, Bio1, Bio6, Bio15, Species, Present)

set.seed(50)
rf <- randomForest(Present~ ., data = Hempson, mtry = best.m, importance=TRUE, ntree = 500)
print(rf)

mtry <- tuneRF(Hempson1[-1],Hempson1$Present, ntreeTry=500,
               stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
print(mtry)
print(best.m)

#Evaluate variable importance
importance(rf)
varImpPlot(rf)
print(rf)


imp <- varImpPlot(rf)
imp <- as.data.frame(imp)
imp$varnames <- rownames(imp)
rownames(imp) <- NULL
imp$var_categ <- rep(1:2, 4)

write.csv(imp, "/Users/kathrynsokolowski/Desktop/meansq.csv")
imp <- read.csv("/Users/kathrynsokolowski/Desktop/meansq1.csv")



imp <- imp %>%
  arrange(IncMSE) %>%
  mutate(varnames=factor(varnames, levels=varnames))

  
p <- ggplot(imp, aes(y=IncMSE, x=varnames)) + 
    geom_segment(aes(xend=varnames, yend=0), size =3, color = "red") +
   geom_point(size= 7, color="red")  +
   ylab("% Mean Square Error") +
  coord_flip() + 
  theme_minimal() +
    labs(title="Variable importance") +
   scale_x_discrete(labels=function(x) gsub("_", " ", x, fixed=TRUE))+
   theme(plot.title=element_text(hjust=0.5), text = element_text(size=30))


jpeg(
  filename="meansq1.jpg",
  width=8,
  height=8,
  units="in",
  res=500)



```


```{r}
summary(rf)

```
Check on GAM as null model
```{r}
n_distinct(Hempson$Species)
as_factor(Hempson)

Hempson$Species <- as.factor(Hempson$Species)

Hempson$Species <- as.numeric(Hempson$Species)
Species <- Hempson$Species 
summary(Species)

class(Hempson$Species)

mod.gam <- gam(Present ~ s(Species) + s(Latitude) +s(Longitude) + s(mean_wc) + s(Bio1) + s(Bio6) + s(Bio15) + s(AridityIndex), method = "GCV.Cp",data = Hempson) 

summary(mod.gam) #model summary


```


Aridity Probabilty of presence
```{r}

# Aridity -----------------------------------------------------------------

min(Hempson$AridityIndex)
max(Hempson$AridityIndex)

Species <- unique(Hempson$Species) #make a list of all Species in the data.frame

predict.wc <- tibble(
  AridityIndex = seq(500, 10000, length = 100)
) #range of environmental variable for prediction

rf.Species.dist <- as_tibble(
  matrix(nrow = nrow(predict.wc), ncol = length(Species))
) #make a blank data.frame to hold the output

#add Species name to each column of the output
colnames(rf.Species.dist) <- Species 

set.seed(49)#ers <-that's for Tyler #set arbitrary seed so results are consistent

for (i in Species) {
  
  suppressWarnings(
    mod.rf <- randomForest(
      Present ~ AridityIndex
      data = subset(Hempson, Species == i),
      replace = TRUE, # bootstrap re-sample each run
      sampsize = 15, # down sample to balance pres/abs
      na.action = na.omit # omit missing data, but could impute: na.roughfix = median or rfimpute = proximity based
    ) 
  )
  
  rf.Species.dist[, i] <- predict(
    mod.rf,                #predict model fit
    newdata = predict.wc,  #across the range of predictor data
    type = 'response'      #output on response scale (0,1)
    #predict.all=TRUE,     #returns list with a fit for each tree, which could be useful later
  )
  
  remove(mod.rf)
  
}

rf.Species.dist <- rf.Species.dist %>%
  as_tibble() %>%
  mutate(
    AridityIndex = predict.wc$AridityIndex
  ) %>%
  pivot_longer(
    -AridityIndex, 
    names_to = "Species", 
    values_to = "Present"
  ) %>%
  arrange(Species) %>%
  select(Species, AridityIndex, Present)

Species_response_plots1 <- 
  ggplot(rf.Species.dist, aes(AridityIndex, Present)) +
  geom_line() +
  facet_wrap_paginate(~Species, ncol = 1, nrow =1, page = 1) +
  theme_bw() 



Species_response_plots1
```

Woody cover response


```{r}
Woody Cover -------------------------------------------------------------
min(Hempson$mean_wc)
max(Hempson$mean_wc)

Species <- unique(Hempson$Species) #make a list of all Species in the data.frame

predict.wc <- tibble(
  mean_wc = seq(0, 1, length = 100)
) #range of environmental variable for prediction

rf.Species.dist <- as_tibble(
  matrix(nrow = nrow(predict.wc), ncol = length(Species))
) #make a blank data.frame to hold the output

#add Species name to each column of the output
colnames(rf.Species.dist) <- Species 

set.seed(49)#ers <-that's for Tyler #set arbitrary seed so results are consistent

for (i in Species) {
  
  suppressWarnings(
    mod.rf <- randomForest(
      Present ~ mean_wc,
      data = subset(Hempson, Species == i),
      replace = TRUE, # bootstrap re-sample each run
      sampsize = 15, # down sample to balance pres/abs
      na.action = na.omit # omit missing data, but could impute: na.roughfix = median or rfimpute = proximity based
    ) 
  )
  
  rf.Species.dist[, i] <- predict(
    mod.rf,                #predict model fit
    newdata = predict.wc,  #across the range of predictor data
    type = 'response'      #output on response scale (0,1)
    #predict.all=TRUE,     #returns list with a fit for each tree, which could be useful later
  )
  
  remove(mod.rf)
  
}

rf.Species.dist <- rf.Species.dist %>%
  as_tibble() %>%
  mutate(
    mean_wc = predict.wc$mean_wc
  ) %>%
  pivot_longer(
    -mean_wc, 
    names_to = "Species", 
    values_to = "Present"
  ) %>%
  arrange(Species) %>%
  select(Species, mean_wc, Present)

Species_response_plots1 <- 
  ggplot(rf.Species.dist, aes(mean_wc, Present)) +
  geom_line() +
  facet_wrap_paginate(~Species, ncol = 3, nrow =4, page = 1) +
  theme_bw() 

Species_response_plots1
```
Precipitation seasonality

```{r}


min(Hempson$Bio15)
max(Hempson$Bio15)

Species <- unique(Hempson$Species) #make a list of all Species in the data.frame

predict.wc <- tibble(
  Bio15 = seq(10, 150, length = 100)
) #range of environmental variable for prediction

rf.Species.dist <- as_tibble(
  matrix(nrow = nrow(predict.wc), ncol = length(Species))
) #make a blank data.frame to hold the output

#add Species name to each column of the output
colnames(rf.Species.dist) <- Species 

set.seed(49)#ers <-that's for Tyler #set arbitrary seed so results are consistent

for (i in Species) {
  
  suppressWarnings(
    mod.rf <- randomForest(
      Present ~ Bio15,
      data = subset(Hempson, Species == i),
      replace = TRUE, # bootstrap re-sample each run
      sampsize = 15, # down sample to balance pres/abs
      na.action = na.omit # omit missing data, but could impute: na.roughfix = median or rfimpute = proximity based
    ) 
  )
  
  rf.Species.dist[, i] <- predict(
    mod.rf,                #predict model fit
    newdata = predict.wc,  #across the range of predictor data
    type = 'response'      #output on response scale (0,1)
    #predict.all=TRUE,     #returns list with a fit for each tree, which could be useful later
  )
  
  remove(mod.rf)
  
}

rf.Species.dist <- rf.Species.dist %>%
  as_tibble() %>%
  mutate(
    Bio15 = predict.wc$Bio15
  ) %>%
  pivot_longer(
    -Bio15, 
    names_to = "Species", 
    values_to = "Present"
  ) %>%
  arrange(Species) %>%
  select(Species, Bio15, Present)


Species_response_plots1 <- 
  ggplot(rf.Species.dist, aes(Bio15, Present)) +
  geom_line() +

  facet_wrap_paginate(~Species, ncol = 1, nrow =1, page = 1) +
  theme_bw() 


Species_response_plots1 <- 
  ggplot(rf.Species.dist, aes(Bio15, Present)) +
  geom_line() +
  facet_wrap_paginate(~Species, ncol = 3, nrow =4, page = 1) +
  theme_bw() 


Species_response_plots1


